---
title: "What factors influence house prices in New York?"
subtitle: "L17G04: ZhenPeng, Callista, Sam, Richard & XiaoMa"
format: 
  revealjs:
    callout-appearance: simple
    slide-number: true
    auto-slide: 20000
    embed-resources: true
    css: style.css
---

# Motivation {.bg-image1}
<!--Slide 2: Motivation-->
**Why is this important?**

:::{.text-container}
- Physical and social factors of an area
- Predict and understand future housing market trends
:::

<!--Slide 3: Dataset Overview-->
# Dataset Overview {.bg-image2}
**Data set:** House Prices in New York `(housing-prices-ge19 (1).txt)`

**Target variable:** Price

```{r, message=FALSE}
# load libraries
library(tidyverse)
library(dplyr)
library(caret)
library(scales)
library(janitor)
library(ggfortify)
library(knitr)
library(kableExtra)

set.seed(123)

raw_data <- read.delim("housing-prices-ge19 (1).txt") %>% clean_names
``` 

```{r, results='hide'}
glimpse(raw_data)
```
<!--decided to drop the variable test, irrelevant-->

```{r, class.source='small'}
data = raw_data %>% select(-test)

```


:::{.smaller}
| Summary       | Count |
|----------------|-------:|
| **Rows**       | 1,734 |
| **Columns**    | 16    |
:::

**Variables:**  

:::{.smaller}
price, lot_size, waterfront, age, land_value, new_construct, central_air, fuel_type, heat_type, sewer_type, living_area, pct_college, bedrooms, fireplaces, bathrooms, rooms
:::


<!--Slide 4: Summary Stat-->
# Summary Statistics {.bg-image3}
```{r}
smry = summary(data$price)
```
Price Summary Statistics

| Min       | 1st Qu    |Median     |Mean       | 3rd Qu    |Max         |   
|-----------|-----------|-----------|-----------|-----------|-----------:|
|5000       |145000     |189700     |211545     |257290     |775000      |

<!--Slide 5: Price Distribution-->
# Price Distribution {.bg-image4}
```{r}
# original histogram
ggplot(data, aes(x = price)) +
  geom_histogram(bins = 80, color = "#d19b75", fill =  "#F4F2EC") +
  labs(x = "Price ($)", y = "Count") +
  scale_x_continuous(labels = scales::comma) +  # show commas instead of scientific notation
  theme_classic(base_size = 10)
```
- right skewed

# Transformations {.bg-image4}
Apply log transformation:

:::{.smaller}
- `ln_Price = log(price)`
- `ln_Lot.Size = log(lot_size)`
- `ln_Land.Value = log(land_value)`


Binary to factors: `new_construct`, `central_air`
:::

```{r}
data <- data %>%
  filter(lot_size != 0)

# apply log transformation
data = data %>%
  mutate(
    ln_Price = log(price),
    ln_Lot.Size = log(lot_size),
    ln_Land.Value = log(land_value),
    new_construct = factor(if_else(new_construct == 1, "Yes", "No")), 
    central_air = factor(if_else(central_air == 1, "Yes", "No"))
  )

# exclude the old variables
ln_data = data %>% select(-price, -lot_size, -land_value)

# new histogram
ggplot(ln_data, aes(x = ln_Price)) +
  geom_histogram(bins = 80, color = "#d19b75", fill =  "#F4F2EC") +
  labs(x = "ln_Price ($)", y = "Count") +
  theme_classic(base_size = 10)


```


<!--Slide 6: Model Building Process-->
# The Process {.bg-image5}
1) Use three variable selection methods to create regression models

:::{.smaller}
  - Forward Selection
  - Backward Elimination
  - Stepwise Selection
  
:::

2) Use cross validation

3) Pick the best model!

```{r}
# Null and Full Model
M0 = lm(ln_Price ~ 1, data = ln_data) # Null model
M1 = lm(ln_Price ~., data = ln_data) # Full model

```


<!--Slide 8: Forward Selection-->
# Model A: Forward Selection {.bg-image6}

```{r}
A = step(M0, direction = "forward", scope = list(lower = M0, upper = M1), trace = F)
```
Formula:

:::{.smaller}
\begin{equation*}
\begin{split}
ln(price) = & \beta_0+\beta_1\cdot\text{living_area}+\beta_2\cdot ln(\text{land_value})+ \beta_3\cdot bathrooms+ \\
& \beta_4\cdot waterfront+ \beta_5\cdot\text{heat_type}+ \beta_6\cdot ln(\text{lot_size}) + \\
& \beta_7\cdot\text{age}+\beta_8\cdot\text{new_construct}+\beta_9\cdot\text{central_air}+ \\
& \beta_{10}\cdot\text{pct_college}+\beta_{11}\cdot\text{rooms}
\end{split}
\end{equation*}
:::

AIC of model A: 678.0835
```{r, results='hide'}
AIC(A)
```


<!--Slide 9: Backward Elimination-->
# Model B: Backward Elimination {.bg-image7}

```{r}
B = step(M1, direction = "backward", trace = F)
```

Formula:

:::{.smaller}
\begin{equation*}
\begin{split}
ln(price)= & \beta_0 + \beta_1\cdot\text{waterfront} + \beta_2\cdot\text{age} + \beta_3\cdot\text{new_construct} + \\
& \beta_4\cdot\text{central_air} + \beta_5\cdot\text{fuel_type} + \\
& \beta_6\cdot\text{living_area} + \beta_7\cdot\text{pct_college} + \\
& \beta_8\cdot\text{bathrooms} + \beta_9\cdot\text{rooms} + \\
& \beta_{10}\cdot ln(\text{lot_size}) + \\
& \beta_{11}\cdot ln(\text{land_value})
\end{split}
\end{equation*}
:::

AIC of model B: 676.1894
```{r, results='hide'}
AIC(B)
```


<!--Slide 10: Stepwise Selection-->
# Model C: Stepwise Selection {.bg-image8 .right-aligned-slide}
```{r}
C = step(M1, direction = "both", trace = F)
```

:::{.left-aligned-text}
Formula:
:::

:::{.smaller}
\begin{equation*}
\begin{split}
ln(price)= & \beta_0 + \beta_1\cdot\text{waterfront} + \beta_2\cdot\text{age} + \beta_3\cdot\text{new_construct} + \\
& \beta_4\cdot\text{central_air} + \beta_5\cdot\text{fuel_type} + \beta_6\cdot\text{living_area} + \\
& \beta_7\cdot\text{pct_college} + \beta_8\cdot\text{bathrooms} + \beta_9\cdot\text{rooms} + \\
& \beta_{10}\cdot ln(\text{lot_size}) + \beta_{11}\cdot ln(\text{land_value})
\end{split}
\end{equation*}
:::

AIC of model C: 676.1894
```{r, results='hide'}
AIC(C)
```


<!--Slide 11: Model Comparison-->
# Model Comparison using Cross Validation {.bg-image9}
K = 10

```{r}
# cross validation
cv = trainControl(method = "cv", number = 10)

cv_A = train(ln_Price ~ ., data = ln_data[, all.vars(formula(A))], method = "lm", trControl = cv)
cv_B = train(ln_Price ~ ., data = ln_data[, all.vars(formula(B))], method = "lm", trControl = cv)
cv_C = train(ln_Price ~ ., data = ln_data[, all.vars(formula(C))], method = "lm", trControl = cv)

aic_list = c(AIC(A), AIC(B), AIC(C))
bic_list = c(BIC(A), BIC(B), BIC(C))

# results table
results = data.frame(
  Model = c("A", "B", "C"),
  RMSE = c(cv_A$results$RMSE, cv_B$results$RMSE, cv_C$results$RMSE),
  Rsquared = c(cv_A$results$Rsquared, cv_B$results$Rsquared, cv_C$results$Rsquared),
  AIC = aic_list,
  BIC = bic_list
)

```



```{r}
# knitr to make table nicer
kable(results, digits = 4) %>%
  kable_styling(full_width = FALSE, position = "center", bootstrap_options = c("striped", "hover"))

```


<!--Slide 12: Choose the Best one-->
# Best Model: Model C {.bg-image10}

:::{.smaller}
\begin{equation*}
\begin{split}
ln(price)= & \beta_0 + \beta_1\cdot\text{waterfront} + \beta_2\cdot\text{age} + \beta_3\cdot\text{new_construct} + \\
& \beta_4\cdot\text{central_air} + \beta_5\cdot\text{fuel_type} + \beta_6\cdot\text{living_area} + \\
& \beta_7\cdot\text{pct_college} + \beta_8\cdot\text{bathrooms} + \beta_9\cdot\text{rooms} + \\
& \beta_{10}\cdot ln(\text{lot_size}) + \beta_{11}\cdot ln(\text{land_value})
\end{split}
\end{equation*}
:::

<!--Slide 13: Residual vs Fitted-->
# Residual vs Fitted {.bg-image11}

::: {layout="[0.75, 0.01, 0.25]"}

```{r}

autoplot(C, which = 1, label.size = 2)

```

- Residuals scattered around 0
- No discernable pattern
:::








<!--Slide 14: QQ Plot-->
# Normal Q-Q Plot {.bg-image12}


::: {layout="[0.75, 0.01, 0.25]"}

```{r}
autoplot(C, which = 2,  label.size = 2)
```

- Majority is normally distributed
- Few outliers at high/low quantiles
:::


<!--Slide 15: Assumption Check-->
# Assumption Check {.bg-image13}

:::{.smaller}
1. Linearity - Yes, mostly linear
2. Normality of Residuals - Yes, mostly normal
3. Homoskedasticity - Yes, fairly consistent
4. Independence - Yes
:::

<!--Slide 16: Interpret Coefficients-->
# Practical Interpretation {.bg-image5}
:::{.smaller}
- **waterfront**: +54% if waterfront
- **age**: -0.1% per year
- **new_construct**: -13% if new construct
- **central_air**: +6% if central air
- **fuel_type**: +9% Gas, +5% Oil, +0.8% Solar, -101% Wood, -43% Other, -28% None
- **living_area**: +0.03%
- **pct_college**: -0.2%
- **bathrooms**: +11% per bathroom
- **rooms**: +0.7% per room
- **ln_Lot.Size**: +11%
- **ln_Land.Value**: +12%
:::

<!--Slide 17: Limitations-->
# Limitations {.bg-image14}
- Spatial dependency
- Possible omitted features


<!--Slide 18: Future Improvements-->
# Future Improvements {.bg-image15}
- Interactions
- Polynomial terms
- LASSO
- Machine Learning Models

<!--Slide 19: Conclusion-->
# Conclusion {.bg-image16}
**Research Question:** What factors influence house prices in New York?

:::{.smaller}
- Location matters
- House characteristics: more bathrooms, larger living area, central air
- Older houses decreases price, but new construct also decreases price
- Fuel type matters
:::

**Overall:** Price is influenced by a combination of location, house feature and neighborhood characteristics, with some factors having much larger effects than others.


# Thank You!







